---
title: "Voter Turnout Analysis"
author: "Christian Lopez"
date: "2025-10-12"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project analyzes county-level socioeconomic variables to understand and predict voter turnout using statistical and machine learning models.

```{r}
df <- read.csv("Data/socioeconomic_voting.csv")
```

```{r}
library(ggplot2)
library(caret)
library(car)
library(betareg)
library(glmulti)
library(randomForest)
library(Metrics)
library(MASS)
library(dplyr)

# Inspect dataset dimensions and variable summaries
dim(df) # 3115 obs, 10 var
summary(df)

# No variables appear to create obvious target leakage.

# Create Census region variable from state names
df1 <- df %>% 
  mutate(Region = case_when(
    State %in% c("CONNECTICUT", "MAINE", "MASSACHUSETTS", "NEW HAMPSHIRE", "RHODE ISLAND",
                 "VERMONT", "NEW JERSEY", "NEW YORK", "PENNSYLVANIA") ~ "Northeast",
    
    State %in% c("ILLINOIS", "INDIANA", "MICHIGAN", "OHIO", "WISCONSIN", "IOWA",
                 "KANSAS", "MINNESOTA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA",
                 "SOUTH DAKOTA") ~ "Midwest",
    
    State %in% c("DELAWARE", "DISTRICT OF COLUMBIA", "FLORIDA", "GEORGIA", "MARYLAND", "NORTH CAROLINA",
                 "SOUTH CAROLINA", "VIRGINIA", "WEST VIRGINIA", "ALABAMA", "KENTUCKY",
                 "MISSISSIPPI", "TENNESSEE", "ARKANSAS", "LOUISIANA", "OKLAHOMA", "TEXAS") ~ "South",
    
    State %in% c("ARIZONA", "COLORADO", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", "UTAH",
                 "WYOMING", "ALASKA", "CALIFORNIA", "HAWAII", "OREGON", "WASHINGTON") ~ "West",
    TRUE ~ NA_character_
  ))

# Check for states that were not assigned a region
sum(is.na(df1$Region))
df1[is.na(df1$Region), ]

# Remove identifying geographic variables not used for modeling
df1 <- df1 %>%
  dplyr::select(-any_of(c("State", "County.Name", "FIPS")))

head(df1) 
```

# EDA

This section performs exploratory data analysis (EDA), including descriptive statistics, outlier detection, and visualization to understand variable distributions and relationships.

[Link 1](https://censusreporter.org/profiles/05000US48301-loving-county-tx/) [Link 2](https://hdpulse.nimhd.nih.gov/data-portal/social/table?age=081&age_options=age25_1&demo=00006&demo_options=education_3&race=00&race_options=race_7&sex=0&sex_options=sexboth_1&socialtopic=020&socialtopic_options=social_6&statefips=48&statefips_options=area_states)

```{r}
# An extreme outlier is observed where bachelor's degree percentage equals 0.
# This corresponds to Loving County, Texas, which has an extremely small population.
# Due to the instability of estimates in very low-population counties, this observation
# is removed to improve model reliability.
dim(df1) # 3115 obs, 8 var
summary(df1) 


df1[df1$Bachelor.s.Degree.or.Higher.Percentage..2018.2022. == 0, ] # obs 2647
df[2647, ] # Loving, Texas

df1 <- df1 %>%
  filter(Bachelor.s.Degree.or.Higher.Percentage..2018.2022. != 0)

dim(df1) # 3114 observations now
df1[df1$Bachelor.s.Degree.or.Higher.Percentage..2018.2022. == 0, ] # is removed

# Correlation
numerical_df <- df1[ ,c("Unemployment.Rate.2020", "County.Median.Household.Income..2021.",
                  "County.Income.Percentile.Within.State..2021.",
                  "Bachelor.s.Degree.or.Higher.Percentage..2018.2022.", "Vote.Percentage")]
cor(numerical_df)
# Strong positive correlations observed between:
# - Median income and income percentile (0.749)
# - Median income and bachelor's degree percentage (0.725)

# Visuals between predictors and response
numerical_cols <- c(
  "Unemployment.Rate.2020",
  "County.Median.Household.Income..2021.",
  "County.Income.Percentile.Within.State..2021.",
  "Bachelor.s.Degree.or.Higher.Percentage..2018.2022.",
  "Vote.Percentage"
)

# Histogram of variables (will try transformations during model building)
lapply(numerical_cols, function(col) {
  hist(df1[[col]],
       main = paste("Histogram of", col),
       xlab = col,
       col = "lightblue",
       border = "darkblue")
})
# Observed distributions:
# - Vote percentage: left-skewed
# - Unemployment rate: right-skewed
# - Median income: right-skewed
# - Bachelor's degree percentage: right-skewed
# - Income percentile: approximately symmetric

plot(density(df1$County.Income.Percentile.Within.State..2021.),
     main = "Density of Income Percentile",
     xlab = "Income Percentile",
     col = "blue", lwd = 2)
# income percentile looks normal

# Identified extreme values in unemployment and income variables.
# Verified that these correspond to valid observations with large populations,
# so they are retained in the dataset.
for (col_name in numerical_cols) {
  boxplot(df1[[col_name]],
          main = paste("Boxplot of", col_name),
          ylab = col_name,
          col = "lightblue",
          border = "darkblue")
}

for (col_name in numerical_cols) {
  x <- df1[[col_name]]
  
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
  outliers <- df1[x < lower_bound | x > upper_bound, ]
  
  print(outliers[, col_name, drop = FALSE])
}

df[173, ] # imperial, california
df[201, ] # san mateo, california

# Shows amount in each group to make sure that none are too small (less than 5%)
categorical_cols <- c("Party", "Region", "Urban.Influence.Code.2013") 

for (col_name in categorical_cols) {
  print(table(df1[[col_name]]))
}
# Urban influence categories are consolidated to ensure sufficient sample size
# within each group for modeling.

# group 1 will now have 1-3 in it, group 2 will have 4-6, ... , group 4 has 10-12
df1 <- df1 %>%
  mutate(UIC_Group = case_when(
    Urban.Influence.Code.2013 %in% 1:3 ~ "1",
    Urban.Influence.Code.2013 %in% 4:6 ~ "2",
    Urban.Influence.Code.2013 %in% 7:9 ~ "3",
    Urban.Influence.Code.2013 %in% 10:12 ~ "4"
  ))

# Check
table(df1$UIC_Group)
# Groups are now big enough
# Remove original UIC column
df1 <- dplyr::select(df1, -Urban.Influence.Code.2013)
```

## Factoring and encoding

```{r}
# Factorization
categorical_cols <- c("Party", "Region", "UIC_Group")
str(df1[categorical_cols])

for (col_name in categorical_cols) {
  df1[[col_name]] <- as.factor(df1[[col_name]])
}

# Confirm categorical variables are factors
str(df1[categorical_cols])

# Data splitting
set.seed(123)
partition <- createDataPartition(df1$Vote.Percentage, p = 0.7, list = FALSE)

train_raw <- df1[partition, ]
test_raw  <- df1[-partition, ]

# Check split balance
dim(train_raw)
dim(test_raw)

mean(train_raw$Vote.Percentage)
mean(test_raw$Vote.Percentage)

range(train_raw$Vote.Percentage)
range(test_raw$Vote.Percentage)

# Check predictor-response relationships 
temp_model <- lm(Vote.Percentage ~ ., data = train_raw)
crPlots(temp_model)
# Component + residual plots indicate approximate linearity between predictors and response.
# No strong nonlinear patterns were observed, so no transformations were applied.

# Recheck influential observations using training data only
plot(temp_model)

# Inspect observations with the largest influence values 
cooks_d <- cooks.distance(temp_model)
top_influential <- order(cooks_d, decreasing = TRUE)[1:3]

train_raw[top_influential, ]

# These observations are retained if they have valid values and reasonable populations.

# Encode categorical variables 
dummy_model <- dummyVars(~ ., data = train_raw[categorical_cols], fullRank = TRUE)

train_dummy <- as.data.frame(predict(dummy_model, newdata = train_raw))
test_dummy  <- as.data.frame(predict(dummy_model, newdata = test_raw))

head(train_dummy)

# Combine numeric variables and encoded categorical variables
numerical_cols <- c(
  "Unemployment.Rate.2020",
  "County.Median.Household.Income..2021.",
  "County.Income.Percentile.Within.State..2021.",
  "Bachelor.s.Degree.or.Higher.Percentage..2018.2022.",
  "Vote.Percentage"
)

train_df <- cbind(train_raw[numerical_cols], train_dummy)
test_df  <- cbind(test_raw[numerical_cols], test_dummy)

dim(train_df) # 2,182 observations
dim(test_df) # 932 observations

```

# Model Building

## Regression

```{r}
# Response is voting percentage (continuous, bounded between 0 and 100).
# A linear model provides interpretability, but predictions may fall outside bounds.
# A beta regression model is more appropriate after scaling the response to (0,1).

# Linear Regression
linear_model <- lm(Vote.Percentage ~ ., data = train_df)
summary(linear_model) # R^2 ≈ 0.186

# Beta Regression

# Scale response variable using TRAIN and TEST 
train_df_beta <- train_df %>%
  mutate(Vote_scaled = Vote.Percentage / 100) %>%
  dplyr::select(-Vote.Percentage)

test_df_beta <- test_df %>%
  mutate(Vote_scaled = Vote.Percentage / 100) %>%
  dplyr::select(-Vote.Percentage)

# Check for boundary issues (beta requires values strictly between 0 and 1)
sum(train_df_beta$Vote_scaled == 0)
sum(train_df_beta$Vote_scaled == 1)
# No exact 0s or 1s detected

# Fit full beta regression model
beta_model <- betareg(Vote_scaled ~ ., data = train_df_beta)
summary(beta_model) # R^2 ≈ 0.189

# AIC values are not directly compared between the linear regression and beta regression models.
# Linear regression is used as a baseline model.
# Beta regression is selected as the main model because voter turnout is continuous
# and bounded, making a proportion-based model more appropriate after scaling.

# Model Diagnostic

# Model diagnostics show residuals mostly centered around zero.
# Some clustering and influential observations are present, but no severe pattern
# suggests that the beta regression model is unusable.
# Residual plots do not show strong evidence that transformations are necessary.

plot(beta_model)
# Diagnostics indicate:
# - Residuals are approximately centered
# - No strong curvature or funneling
# - Some high-leverage points, but no critical issues

# Check residual relationships with predictors
resid_q <- residuals(beta_model, type = "quantile")

plot(train_df_beta$Unemployment.Rate.2020, resid_q,
     xlab = "Unemployment Rate (2020)",
     ylab = "Quantile Residuals")

plot(train_df_beta$County.Median.Household.Income..2021., resid_q,
     xlab = "Median Household Income (2021)",
     ylab = "Quantile Residuals")

plot(train_df_beta$County.Income.Percentile.Within.State..2021., resid_q,
     xlab = "Income Percentile",
     ylab = "Quantile Residuals")

plot(train_df_beta$Bachelor.s.Degree.or.Higher.Percentage..2018.2022., resid_q,
     xlab = "Bachelor's Degree (%)",
     ylab = "Quantile Residuals")

# Categorical checks (already encoded)
# No strong pattern differences observed

# No transformations are required.

# Model Selection

# Selected model based on AIC comparison
beta_model_select <- betareg(
  Vote_scaled ~ 
    Unemployment.Rate.2020 +
    County.Median.Household.Income..2021. +
    County.Income.Percentile.Within.State..2021. +
    Bachelor.s.Degree.or.Higher.Percentage..2018.2022. +
    Party.REPUBLICAN + Region.South +
    UIC_Group.2 + UIC_Group.3 + UIC_Group.4,
  data = train_df_beta
)

# Compare full and reduced beta models
AIC(beta_model)
AIC(beta_model_select)

summary(beta_model_select)

# The full beta model has a slightly lower AIC than the reduced model.
# However, the AIC difference is small, and the reduced model has fewer predictors.
# The reduced model is selected for simplicity and interpretability while maintaining
# similar model fit.

# Multicollinearity Check

vif_model <- lm(
  Vote_scaled ~ 
    Unemployment.Rate.2020 +
    County.Median.Household.Income..2021. +
    County.Income.Percentile.Within.State..2021. +
    Bachelor.s.Degree.or.Higher.Percentage..2018.2022. +
    Party.REPUBLICAN + Region.South +
    UIC_Group.2 + UIC_Group.3 + UIC_Group.4,
  data = train_df_beta
)

vif(vif_model)

# All VIF values are below 5, indicating no serious multicollinearity issues.

# Final model
final_model <- beta_model_select
```

## Random Forest

Random forest is included as a flexible tree-based model. It may capture nonlinear patterns and interactions better than regression models, but it is less directly interpretable.

```{r}
set.seed(123)

rf_model <- randomForest(
  Vote.Percentage ~ .,
  data = train_df,
  ntree = 1000,
  mtry = floor(sqrt(ncol(train_df) - 1)),
  importance = TRUE
)

rf_model
```

# Model Evaluation

## Regression Evaluation

```{r}
# Beta regression predictions on test data
beta_preds <- predict(final_model, newdata = test_df_beta, type = "response")

# Test errors
rmse_beta <- rmse(test_df_beta$Vote_scaled, beta_preds)
mae_beta  <- mae(test_df_beta$Vote_scaled, beta_preds)

rmse_beta
mae_beta

# RMSE = 0.139 and MAE = 0.112 on the scaled response.
# This means predictions are off by about 11.2 percentage points on average.

# Test R-squared
y_actual <- test_df_beta$Vote_scaled

rss <- sum((y_actual - beta_preds)^2)
tss <- sum((y_actual - mean(y_actual))^2)

test_r2 <- 1 - rss / tss
test_r2

# Test R-squared is about 0.190, meaning the model explains about 19% of the
# variation in test-set turnout using the included predictors.

# Predicted vs actual
plot(y_actual, beta_preds,
     xlab = "Actual Turnout (scaled)",
     ylab = "Predicted Turnout (scaled)",
     main = "Beta Regression: Predicted vs Actual")

abline(0, 1, col = "red")

# Final model summary
summary(final_model)

# Coefficients are interpreted on the logit scale because beta regression uses
# a logit link for the mean model.
# Example:
# Region.South has a negative coefficient, meaning counties in the South are
# associated with lower expected turnout compared to the reference region,
# holding other variables constant.
```

## Random Forest Evaluation

```{r}
rf_preds <- predict(rf_model, newdata = test_df)

rf_rmse <- RMSE(rf_preds, test_df$Vote.Percentage)
rf_mae  <- MAE(rf_preds, test_df$Vote.Percentage)

rf_rmse
rf_mae

# RMSE is approximately 12.81 percentage points.
# MAE is approximately 9.99 percentage points.
# These errors are lower than the beta regression errors after converting
# beta regression errors back to percentage-point scale.

# Random Forest test R-squared
y_actual_rf <- test_df$Vote.Percentage

rss_rf <- sum((y_actual_rf - rf_preds)^2)
tss_rf <- sum((y_actual_rf - mean(y_actual_rf))^2)

rf_r2_test <- 1 - rss_rf / tss_rf
rf_r2_test

# Test R-squared is approximately 0.308, meaning the random forest explains
# about 30.8% of variation in test-set turnout.

# Variable importance
varImpPlot(rf_model)
importance(rf_model)

# Based on %IncMSE, the most important predictors are:
# bachelor's degree percentage, Republican vote indicator, unemployment rate,
# South region indicator, and median household income.

# Overall, the random forest had stronger predictive performance than beta regression,
# with lower RMSE/MAE and higher test R-squared.
# The beta regression model remains useful for interpretation, while the random forest
# is better for prediction.
```
